---
title: "MA461 Assignment 3"
author: "Conor O'Donoghue"
date: "April 21, 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Creating the Function
Our goal is to calculate the genetic distance of two sequences using the 
Kimura 2-parameter model, but this model utilizes two parameters: time, and 
the ratio of transition rate ($\alpha$) to transversion rate ($\beta$) that we don't have.

So, our goal is to do a maximum likelihood estimate of the genetic distance,
using the optimx package to maximize the probability of the observed sequences
with time and the ratio as free parameters.

## Why the ratio of a to b and not a and b as separate parameters?
As part of scaling the generator matrix, we need to set $2\beta+\alpha=1$. 
With this constraint, $\alpha$ and $\beta$ can both be calculated from their ratio instead.
Plugging $\alpha=1-2\beta$ into $\frac{\alpha}{\beta}=r$ gives
$\frac{1-2\beta}{\beta}=r$, which rearranges to $\beta=\frac{1}{r+2}$.
Thus we can calculate $\beta$ from r, and obtain $\alpha$ from $\alpha=1-2\beta$.

```{r, message=FALSE}
library(expm)
library(optimx)

# Functions for converting sequences to vector of integers
nuc2int <- function(nuc){
  nucs = c("A","C","G","T")
  nuc_int <- match(nuc, nucs)
  return(nuc_int)
}
seq2int <- function(seq){
  ints <- sapply(strsplit(seq,""), nuc2int)
  return(ints)
}

# kimura_2p function
# t = time, r = ratio of alpha/beta, s1=sequence 1, s2=sequence 2
kimura_2p <- function(params, s1, s2) {
  t=params[1]
  r=params[2]
  # Calculating scaled a and b from r
  b = 1/(r+2)
  a = 1 - 2*b
  # Filling the generator matrix
  # It's not very large, so I find it easiest to just type it out.
  G = array(c(-1,b,a,b,b,-1,b,a,a,b,-1,b,b,a,b,-1), c(4,4))
  # Obtaining the probability matrix using t and G
  Pt = expm(t*G)
  s1_ints = seq2int(s1)
  s2_ints = seq2int(s2)
  prob = Pt[s1_ints[1],s2_ints[1]]
  for(i in 2:length(s1_ints)){
    s1_int <- s1_ints[i]
    s2_int <- s2_ints[i]
    prob <- prob * Pt[s1_int, s2_int]
  }
  return(prob)
}
```
# Optimizing time and ratio

```{r}
seq1 = "AGTCCATGAT"
seq2 = "ACGTCGTGCT"
optimx(par=c(1,2),
       kimura_2p,
       s1=seq1, s2=seq2,
       control=list(maximize=T))
```

It looks like BFGS found a local maximum close to the starting intial value r=2,
but Nelder-Mead found a slightly higher value at r=1.63.
Both of their estimations of t were close, 
but Nelder-Mead found a higher probability with t=0.831.

This means that the maximum likelihood estimate of the genetic distance - the number of substitutions per site since the sequences diverged from their common ancestor - is 0.83.

# Test against distance formula
```{r}
# 2 transitions out of 10 sites
p = 2/10
# 3 transversions out of 10 sites
q = 3/10
d = -0.5*log((1-2*p-q)*sqrt(1-2*q))
d
```

Using the formula provided, we obtain the same answer for genetic distance: 0.831. This corroborates our result as an accurate estimation of genetic distance.
